---
title: "Data Analysis:National Basketball Association"
subtitle: "Recommendation of Chicago Bulls 5 best players"
date: ''
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: '3'
  html_document:
    fig_caption: yes
    number_sections: yes
    self_contained: yes
    theme: flatly
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE,echo=F)
```

# Introduction

## Background information 

Basketball is a team sport in which two teams, usually each with five players, compete against one another on a rectangular court with the main goal of preventing the opposing team from shooting through their own hoop while putting the ball through the defender's hoop, which is a basket with a diameter of 18 inches (46 cm) mounted 10 feet (3.048 m) high to a backboard at each end of the court. Except when made from behind the three-point line, a field goal is worth three points(Roe, 2009). Timed play is stopped following a foul, and the player who committed the foul or who was chosen to attempt a technical foul is awarded one, two, or three one-point free throws.The team with the most points at the conclusion of regulation play wins, but if the score is tied when regulation play ends, extra time (overtime) must be played(Roe, 2009).

Players move the ball forward by either passing it to a teammate or bouncing it while running or walking (dribbling), both of which require a high level of ability. Players can employ a variety of shots on offense, such as layups, jump shots, and dunks. On defense, they can steal the ball from a player who is dribbling, intercept passes, or block shots(Roe, 2009). Finally, either the offense or the defense can grab rebounds, which are missed shots that rebound off the rim or backboard. It is against the rules to carry the ball, elevate or drag one's pivot foot when not dribbling, or hold the ball in both hands before starting to dribble again(Roe, 2009).

There are five playing positions for each side's five players. The center is typically the tallest player, the power forward is second tallest and strongest, the small forward is slightly shorter but more agile, and the shooting guard and point guard are the shortest players or best ball handlers. The point guard executes the coach's game plan by managing the execution of offensive and defensive plays (player positioning). Players can play one-on-one, two-on-two, and three-on-three informally(Radu & Nini, 2018).

Basketball was created in 1891 by Canadian-American gym instructor James Naismith in Springfield, Massachusetts, and has since grown to be one of the most well-liked and watched sports in the entire globe. The National Basketball Association (NBA), which recruits the majority of its talent from American college basketball, is the most significant professional basketball league in terms of fandom, earnings, talent, and level of competition(Radu & Nini, 2018).

The top teams from national leagues outside of North America are eligible to compete in continental competitions like the EuroLeague and the Basketball Champions League Americas. The sport's two biggest international competitions, the Men's Olympic Basketball Tournament and the FIBA Basketball World Cup, draw the best national teams from all over the world. Regional competitions for national teams, such as EuroBasket and the FIBA AmeriCup, are held on each continent(Radu & Nini, 2018).

Top national teams from continental championships compete in the Women's Olympic Basketball Tournament and FIBA Women's Basketball World Cup. The WNBA, along with the popular NCAA Women's Division I Basketball Championship, is the primary North American league, while the most powerful European clubs play in the EuroLeague Women(Taylor, 2012).


## Description of the scenario 

Chicago Bulls is one of the thirty teams which compete in National Basketball Association.In the most recent NBA season (2018-19 as per study), the team was placed 27th out of 30 (based on win-loss record).The team's budget for player contracts next season(2019-20 as per study) is $118 million, ranked 26th out of 30.If the team is in position 26 in terms of budget,we expect the team's performance be 26 or any position between 1 and 26 but not 27.

## The aim of the project 

This study is aimed at finding the best five starting players for Chicago Bulls basket ball team across all the 30 teams putting into consideration various constraints such as team financial constraint(budget).The point guard (PG), shooting guard (SG), small forward (SF), power forward (PF), and center (C) are the five positions in basketball which are required to be filled at start. On the floor, each position has a unique set of responsibilities, and the best basketball teams have incredible harmony between each position.

In order to choose these 5 players,it is crucial to understand whether the player's performance depends on the salary the team pays the player,the player's age and even the position the player plays. With this established,it would be quite easy to select and recommend the players.

## Justification and importance 

In the most recent NBA season (2018-19), Chicago Bulls basket ball team was placed 27th out of 30 (based on win-loss record).The failure or success of a team depends on its players alongside with others such as coach and key advisers. With an attempt to improve its performance,it was necessary to evaluate its players and search out across the thirty teams,the best five starting players.Poor performance of the team justices the importance of the study to statistically analyse and recommend the best five starting players putting into considerations the team financial constraint. 

# Reading and cleaning the raw data 

The study uses **NBA player statistics**,**NBA player salaries** data set in order to achieve its aforementioned objective. These data sets were read into the R studio software by use of **read.csv()** function since the files were already stored in CSV format.A text file with a specific format known as a CSV (comma-separated values) file allows data to be saved in a table-structured format.

The data sets were then joined together into 1 data set using **merge()** function which is in multi-functional **tidyverse** R package.  
  
Using **select()** function,the study selected the only 6 variables or column which were of interest. These columns include;
* player_name-This columns shows the names of various players within the thirty national teams.
* Pos-This column shows the position in the field which a given selected player plays
* Age-This column shows the age of the player in terms of years
* Tm-This column shows the team in which the given player plays
* PTS-This column shows the performance of the player in terms of point given
* Salary-This column shows the salary earned by a given player

```{r}
library(tidyverse)
library(jtools)
library(kableExtra)
library(ggplot2)
Player<-read.csv("2018-19_nba_player-statistics.csv")
payroll<-read.csv("2019-20_nba_team-payroll.csv")
payroll[]<-lapply(payroll,gsub,pattern="$",fixed=TRUE,replacement="")
payroll$salary<- as.numeric(gsub(",","",payroll$salary))

playersalaries<-read.csv("2018-19_nba_player-salaries.csv")

player<-merge(Player,playersalaries,all = T)
player<-player%>%select(player_name,Pos,Age,Tm,PTS,salary)
head(player,7)%>%
 kbl(caption = "Portion of the data") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")

```


# Exploratory analysis 

## Checking for errors and missing values within the datasets 

```{r}
missing=sapply(player, function(x) sum(is.na(x)))%>%data.frame()
names(missing)<-"NA"
missing%>%
 kbl(caption = "Missing") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")

```

There are a total of 278 missing values in the data set to be used. There is no any missing value in the player name,64 missing values in the pos column,64 missing values in Age,64 missing values in Tm,64 missing values in PTS and 22 missing values in the salary column.This means that 36% of our data has missing values. We therefore can not delete these values but instead deal with them through mean imputation. The columns were already in correct data type and therefore no data types errors detected.

```{r}
player$Age[is.na(player$Age)] <- mean(player$Age,na.rm = TRUE)
player$Tm[is.na(player$Tm)] <- mean(player$Tm,na.rm = TRUE)
player$PTS[is.na(player$PTS)] <- mean(player$PTS,na.rm = TRUE)
player$salary[is.na(player$salary)] <- mean(player$salary,na.rm = TRUE)
player<-na.omit(player)
```


## Checking the distribution of variables 

### Distribution of players' age
```{r, out.width="70%"}
attach(player)
par(mfrow=c(1,2))
hist(Age)
boxplot(Age)
par(mfrow=c(1,1))
```

players' age seems to be almost normally distributed but slightly skewed to the right. Only two variables are considered as extreme values of age.

### Distribution of players' points

```{r, out.width="70%"}
par(mfrow=c(1,2))
hist(PTS,main ="Histogram of points",xlab = "Player points" )
boxplot(PTS)
par(mfrow=c(1,1))
```

The players' points are not normally distributed. They are skewed to right.There seems to have several outliers(extreme values) just above the maximum value.

### Distribution of players' salary
```{r, out.width="70%"}
attach(player)
par(mfrow=c(1,2))
hist(salary)
boxplot(salary)
par(mfrow=c(1,1))
```

The players' salary is not normally distributed. They are skewed to right.There seems to have several outliers(extreme values) just above the maximum value.

## Checking for relationships between variables

### Relationship between players' points and salary

```{r, out.width="70%"}
ggplot(player, aes(x=salary, y=PTS)) + 
  geom_point(color="#E69F00")+
  geom_smooth(method=lm, se=FALSE)+ theme(plot.title = element_text(hjust = 0.5))+ylab("Perfomance in points")+xlab("Salary") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),legend.position="none")+ labs(title = "Relationship between players' points and salary")
```


The scatter plot shows that there is a positive direct relationship between the players' performance in terms of points and salary. An increase in salary is associated with an increase in players' points and the opposite is true. 

### Relationship between players' points and Age

```{r, out.width="70%"}
ggplot(player, aes(x=Age, y=PTS)) + 
  geom_point(color="#E69F00")+
  geom_smooth(method=lm, se=FALSE)+ theme(plot.title = element_text(hjust = 0.5))+ylab("Perfomance in points")+xlab("Age") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),legend.position="none")+ labs(title = "Relationship between players' points and age")
```

There seems to have an inverse and negative relationship between age and performance in terms of points. This means that an increase in age is associated with a decrease in performance.

### Relationship between age and positions

```{r, out.width="70%"}
pp=aggregate(cbind(Age)~Pos,FUN=mean,player)
ggplot(pp, aes(x = reorder(Pos, -Age), y = Age))+ geom_bar(stat="identity")+ theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=45, hjust=1))+ylab("Age")+xlab("Field position") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),legend.position="none")+ labs(title = "Age vs field position")
```

SG-PF seems to be the field position with the aged players and C-PF seems to be the youngest.

### Relationship between salary and positions

```{r, out.width="70%"}
pp=aggregate(cbind(salary)~Pos,FUN=mean,player)
ggplot(pp, aes(x = reorder(Pos, -salary), y = salary))+ geom_bar(stat="identity")+ theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=45, hjust=1))+ylab("Salary")+xlab("Field position") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),legend.position="none")+ labs(title = "Salary vs field position")
```

SF-SG seems to be the position paid the highest salary amount and C-PF with the least salary amount.

### Relationship between perfomance(points) and positions

```{r, out.width="70%"}
pp=aggregate(cbind(PTS)~Pos,FUN=mean,player)
ggplot(pp, aes(x = reorder(Pos, -PTS), y = PTS))+ geom_bar(stat="identity")+ theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=45, hjust=1))+ylab("Perfomance in points")+xlab("Field position") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),legend.position="none")+ labs(title = "Perfomance in points vs field position")
```

SF-SG seems to be the best performing position and PF-C is the least performing position.

## Justification for decisions made about data modelling 

To understand whether the player's performance depends on the salary the team pays the player,the player's age and even the position the player plays,we need to conduct a multiple linear regression. This is because the modelling involves more than 1 independent variable and one dependent variable. 

By fitting a line to the observed data, regression models are used to describe relationships between variables. You can estimate a dependent variable's change as an independent variable or set of independent variables changes using regression.

One dependent variable and two or more independent variables are estimated using multiple linear regression. If you want to know how strongly a link exists between two or more independent variables and one dependent variable, as well as how much the dependent variable changes when the independent variables change at a particular level, you can use multiple linear regression.

# Data modelling and results 

## Data modelling 

```{r}
model<-lm(PTS~salary+Age+Pos,data=player)
summary(model)
```


## Assumption checking 

```{r}
par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))
```


### Linear relationship assumption

We will use **_Residuals vs Fitted_** plot to check this assumption.
The residuals "bounce" about the 0 line at random. This shows that it is plausible to assume that the relationship is linear. Around the 0 line, the residuals generally form a "horizontal band." This suggests that the error terms' variances are equivalent. From the fundamentally random pattern of residuals, no one residual "sticks out." There might not be any outliers, according to this.

### Normal distribution of errors assumption

We will use **_Normal Q-Q_** plot to check this assumption.
Given that the points are mostly within 45 degrees of the reference line, the data is normally 
distributed.

#### Homoskedasticity

We will use **_Scale-Location_** to check homogeneity of variance of the residuals. We can confirm that the red line generally follows the plot's horizontal axis. Therefore, it is likely that 
our regression model will satisfy the homoscedasticity assumption. In other words, at all fitted values,the dispersion of the residuals is roughly equal. Additionally, we can confirm that there isn't any 
discernible pattern in the residuals. In other words, the residuals should have nearly similar variability at all fitted values and be randomly distributed around the red line, which is what happens in our situation.


## Model output and interpretation 

```{r}
summary(model)

plot_summs(model)

```

According to the mode output we have F(12,695) = 27.78, p = 0.00. With small p-value,this model is statistically significant. From the model,only salary and age of a player significantly predicts the players performance in terms of points. This is because of their corresponding p-value which are less than 0.05 at 95% confidence interval and 5% level of acceptance.

With a positive estimate for salary(3.664e-05),the average player's performance in terms of points increases by 3.664e-05 for every unit increase in salary holding all other factors constant.With a negative estimate for salary(-2.175e+01),the average player's performance in terms of points decreases by 2.175e+01 for every unit increase in age holding all other factors constant.

# Player recommendations 

From the model output and interpretation,it is very clear that players who are well paid tend to perform better. In addition,it is very clear that young players are expected to perform better as compared to old players. Therefore,the player selection will be based on salary(restricting ourselves to budget below $118 million).

We will recommend one player per each position for the five major positions;point guard (PG), shooting guard (SG), small forward (SF), power forward (PF), and center (C). For each position,we will first select the top 5 players in terms of performance(measured in points). Out of the top 5 players,we will take the youngest best performing. 

## Player for PF
```{r}
PF=subset(player,player$Pos=="PF")
PF=head(PF%>%arrange(-PTS),5)
PF%>%
 kbl(caption = "Top 5 PF players") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")
```


Giannis Antetokounmpo(from MIL team) is the youngest best PF player. The player is 24 years old and is paid a salary of $24157304.

## Player for C
```{r}
C=subset(player,player$Pos=="C")
C=head(C%>%arrange(-PTS),5)
C%>%
 kbl(caption = "Top 5 C players") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")
```

Karl-Anthony Towns(from MIN team) is the youngest best center player who is aged 23 years old.The player is paid a salary of $7839435.

## Player for PG
```{r}
PG=subset(player,player$Pos=="PG")
PG=head(PG%>%arrange(-PTS),5)
PG%>%
 kbl(caption = "Top 5 PG players") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")
```

D'Angelo Russell(from BRK team) is the youngest best PG player. The player is 22 years old and is paid a salary of $7019698.

## Player for SF

```{r}
SF=subset(player,player$Pos=="SF")
SF=head(SF%>%arrange(-PTS),5)
SF%>%
 kbl(caption = "Top 5 SF players") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")
```

Kawhi Leonard(from TOR team) is the youngest best PF player. The player is 27 years old and is paid a salary of $23114066.

## Player for SG

```{r}
SG=subset(player,player$Pos=="SG")
SG=head(SG%>%arrange(-PTS),5)
SG%>%
 kbl(caption = "Top 5 SG players") %>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = "hold_position")
```

Donovan Mitchell(from UTA team) is the youngest best PF player. The player is 22 years old and is paid a salary of $3111480.

The total budget for the 5 players is $ 65241983. This is just 55.29% of the total proposed budget of %118 million.

# Summary 

The study investigated the key factors which should be considered when selecting the best 5 players apart from their points earned(performance). Using performance alone,can bring the best 5 players but the total budget can surpluses the allocated budget of $118 million. The study found that age and salary paid to a player significantly affect the player's performance.

5 players were selected best on the 5 major positions in basket ball. For each position,we selected the top 5 players in terms of performance(measured in points). Out of the top 5 players,we took the youngest best performing player. Therefore the 5 reccommnded players are;

* PF-Giannis Antetokounmpo(from MIL team) is the youngest best PF player. The player is 24 years old and is paid a salary of $24157304.
* Center-Karl-Anthony Towns(from MIN team) is the youngest best center player who is aged 23 years old.The player is paid a salary of $7839435.
* PG-D'Angelo Russell(from BRK team) is the youngest best PG player. The player is 22 years old and is paid a salary of $7019698.
* SF-Kawhi Leonard(from TOR team) is the youngest best PF player. The player is 27 years old and is paid a salary of $23114066.
* SG-Donovan Mitchell(from UTA team) is the youngest best PF player. The player is 22 years old and is paid a salary of $3111480.

# Reference

Radu, A., &amp; Nini, F. (2018). Women’s basketball. THE Science of Basketball, 127–142. https://doi.org/10.4324/9781315204000-6 

Roe, D. (2009). Basketball. African American Studies Center. https://doi.org/10.1093/acref/9780195301731.013.45253 

Taylor, L. (2012). Competing risks in basketball … competing risks in basketball … competing risks in basketball …. CHANCE, 25(2), 31–36. https://doi.org/10.1080/09332480.2012.685367 

